---
title: "QC and Preprocessing of 16S rRNA Sequencing Data"
---

# Load required packages

```{r}
library(phyloseq)
library(tidyverse)
library(glue)
library(vegan)
library(qiime2R)
library(decontam)
library(microbiome)
library(ggrepel)
library(patchwork)
library(rstatix)
library(ggtext)
```

# Load files

```{r}
setwd("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Microbiome Analysis")
metadata <- read_csv("Metadata Processing/metadata_final.csv")
```

Remove samples with missing ddPCR values

```{r}
metadata <- drop_na(metadata, ddPCR)
```

Load QIIME2 artefacts into phyloseq object

```{r}
physeq <- qza_to_phyloseq(features = "QIIME2 Output/filtered-metadata-table.qza", # otu_table
                          tree = "QIIME2 Output/merged-midrooted-tree.qza",
                          taxonomy = "QIIME2 Output/taxonomy-merged-final.qza")
physeq
```

# QC

## Verify concordance between qza files and metadata

```{r}
gplots::venn(list(metadata = metadata$sample_id, physeq = sample_names(physeq)))
```

42 samples are missing ddPCR values and were removed from metadata.

Identify these samples

```{r}
setdiff(sample_names(physeq), metadata$sample_id)
```

Assign metadata as the sample_data slot of physeq

```{r}
metadata <- column_to_rownames(metadata, var = "sample_id")
sample_data(physeq) <- metadata
summarize_phyloseq(physeq)
```

```{r}
sample_data_df <- meta(physeq) # extract sample_data as a dataframe
table(sample_data_df$diagnosis)
```

## Plot library size per sample

```{r}
sample_data_df <- sample_data_df %>% 
  mutate(library_size = sample_sums(physeq)) %>% # total no. of reads for each sample (sums across all taxa in otu_table)
  arrange(library_size) %>% # sort by library size
  mutate(index = seq(nrow(sample_data_df)))
sample_data_df %>%
  ggplot(aes(x = index, y = library_size, color = sample_type)) +
  geom_point(size = 0.5)
```

```{r}
sample_data_df %>%
  ggplot(aes(x = ddPCR, y = library_size, color = sample_type)) +
  geom_point(size = 0.5)
```

## Identify contaminants using two methods

Frequency: Contaminants should vary inversely with no. of 16S gene copies (ddPCR values)

Prevalence: Contaminants should be more prevalent in negative controls

Identify negative controls for prevalence-based contaminant identification

```{r}
physeq_nr <- subset_samples(physeq, sample_type != "Water Oral Control") # remove oral rinse controls
sample_data(physeq_nr)$is_ctrl <- sample_data(physeq_nr)$sample_type != "BAL"
```

### Combined method: Accounts for both frequency and prevalence

We usually use this method for robustness

```{r}
comb_contam <- isContaminant(physeq_nr, method = "combined", conc = "ddPCR", neg = "is_ctrl") # returns a data frame
```

```{r}
table(comb_contam$contaminant)
```

```{r}
head(which(comb_contam$contaminant)) # where in comb_contam to look
```

Plot frequency against DNA concentration of 20 random contaminants. Frequency of contaminants should vary inversely with concentration.

```{r}
comb_contam_sample <- taxa_names(physeq_nr)[sample(which(comb_contam$contaminant), 20)]
plot_frequency(physeq_nr, comb_contam_sample, conc = "ddPCR") + xlab("DNA Concentration (gene/mL)")
```

Plot presence/absence of contaminants in samples vs controls. Contaminants should be of low prevalence in true samples.

```{r}
physeq_pa <- transform_sample_counts(physeq_nr, function(abund) 1 * (abund > 0)) # transform abundance counts in otu_table to presence/absence binary
physeq_pa_spl <- subset_samples(physeq_pa, !is.na(diagnosis))
physeq_pa_ctrl <- subset_samples(physeq_pa, is.na(diagnosis))
df_pa <- data.frame(samples = taxa_sums(physeq_pa_spl), # prevalence of each taxa in true samples and negative controls
                    controls = taxa_sums(physeq_pa_ctrl),
                    is_contam = comb_contam$contaminant)
df_pa %>% 
  ggplot(aes(x = controls, y = samples, color = is_contam)) +
  geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)") 
```

The 9 identified contaminants clustered together with true taxa may not be true contaminants.

Plot contaminant abundance across samples

```{r}
physeq_contam <- prune_taxa(comb_contam$contaminant, physeq_nr) # only subset contaminants from physeq
sample_order <- names(sort(sample_sums(physeq_contam))) # order by total abundance
plot_bar(physeq_contam, fill = "is_ctrl") + 
  scale_x_discrete(limits = sample_order)
```

Contaminants are generally more abundant in controls. 6 samples may be contaminated.

Identify highly abundant contaminants (1000 or more reads in at least 1 sample)

```{r}
is_abund_contam <- genefilter_sample(physeq_contam, filterfun_sample(function(reads) reads >= 1000))
physeq_abund_contam <- prune_taxa(is_abund_contam, physeq_contam)
```

```{r}
tax_table_abund_contam <- as.data.frame(tax_table(physeq_abund_contam))
```

### Prevalence method: Only accounts for prevalence

```{r}
prev_contam <- isContaminant(physeq_nr, method = "prevalence", neg = "is_ctrl")
table(prev_contam$contaminant)
```

```{r}
head(which(prev_contam$contaminant))
```

### Export contaminants obtained from both methods for manual inspection

```{r}
tax_table_mat <- as.matrix(tax_table(physeq_nr))
```

```{r}
write.table(tax_table_mat[which(comb_contam$contaminant), ], file = "QC/comb_contaminants.txt", col.names = NA, sep = "\t") # col.names = NA adds blank column name for column of row (taxa) names
write.table(tax_table_mat[which(prev_contam$contaminant), ], file = "QC/prev_contaminants.txt", col.names = NA, sep = "\t")
```

## How many taxa are not present in any sample?

```{r}
sum(taxa_sums(physeq_nr) == 0)
```

## Examine inter- and intra-group differences using PCoA plot

### Using weighted UniFrac distance on full dataset

```{r}
ord <- ordinate(physeq, method = "PCoA", distance = "wunifrac")
pcoa <- plot_ordination(physeq, ord, color = "sample_type", shape = "study", title = "PCoA using weighted UniFrac distance") +
  geom_point(size = 0.5) +
  scale_colour_manual(values = c("#FAAB18", "#1380A1", "#990000", "#588300", "#778899")) +
  theme(plot.title = element_text(hjust = 0.5))
pcoa
```

Draw confidence ellipses

```{r}
pcoa + stat_ellipse(aes(group = sample_type, color = sample_type)) +
  geom_text_repel(data = filter(pcoa$data, sample_type == "BAL"),
                  aes(label = rownames(filter(pcoa$data, sample_type == "BAL"))),
                  max.overlaps = 25, color = "black", size = 3) +
  theme_classic() + theme(plot.title = element_text(hjust = 0.5))
```

There are intra-group differences within the BAL samples.

### Remove oral rinse controls

```{r}
ord_nr <- ordinate(physeq_nr, method = "PCoA", distance = "wunifrac")
pcoa_nr <- plot_ordination(physeq_nr, ord_nr, color = "sample_type", shape = "study", title = "PCoA excluding oral rinse controls") +
  geom_point(size = 0.5) +
  scale_colour_manual(values = c("#FAAB18", "#1380A1", "#990000", "#588300")) +
  theme(plot.title = element_text(hjust = 0.5))
pcoa_nr + stat_ellipse(aes(group = sample_type, color = sample_type)) +
  geom_text_repel(data = filter(pcoa_nr$data, sample_type == "BAL"),
                  aes(label = rownames(filter(pcoa_nr$data, sample_type == "BAL"))),
                  max.overlaps = 25, color = "black", size = 3) +
  theme_classic() + theme(plot.title = element_text(hjust = 0.5))
```

### Using Bray-Curtis distance

```{r}
ord_bray <- ordinate(physeq, method = "PCoA", distance = "bray")
pcoa_bray <- plot_ordination(physeq, ord_bray, color = "sample_type", shape = "study", title = "PCoA using Bray-Curtis distance") +
  geom_point(size = 0.5) +
  scale_colour_manual(values = c("#FAAB18", "#1380A1", "#990000", "#588300", "#778899")) +
  theme(plot.title = element_text(hjust = 0.5))
pcoa_bray
```

PCoA components explain a smaller % of variance.

### Remove contaminant taxa

```{r}
physeq_nc <- prune_taxa(!comb_contam$contaminant, physeq_nr)

ord_nc <- ordinate(physeq_nc, method = "PCoA", distance = "wunifrac")
pcoa_nc <- plot_ordination(physeq_nc, ord_nc, color = "sample_type", shape = "study", title = "PCoA excluding contaminant taxa") +
  geom_point(size = 0.5) +
  scale_colour_manual(values = c("#FAAB18", "#1380A1", "#990000", "#588300", "#778899"))
pcoa_nc + stat_ellipse(aes(group = sample_type, color = sample_type)) +
  geom_text_repel(data = filter(pcoa_nc$data, sample_type == "BAL"),
                  aes(label = rownames(filter(pcoa_nc$data, sample_type == "BAL"))),
                  max.overlaps = 25, color = "black", size = 3) +
  theme_classic() + theme(plot.title = element_text(hjust = 0.5))
```

Fewer true samples overlap with controls. We will remove contaminant taxa from subsequent analyses.

Sequencing statistics

```{r}
# Original dataset
physeq_nr
summarize_phyloseq(physeq_nr)

# After removing contaminant taxa
physeq_nc
summarize_phyloseq(physeq_nc)
```

### Dissertation figure

```{r}
theme_set(theme_bw() + theme(text = element_text(family = "Arial"),
                             legend.title = element_text(face = "bold"),
                             plot.tag = element_text(face = "bold")))

physeq_nr <- subset_samples(physeq_nr, diagnosis == "IPF" | is_ctrl) # only keep IPF and control samples for visualisation
sample_data(physeq_nr)$study <- factor(sample_data(physeq_nr)$study,
                                       levels = c("BRU", "PROFILE", "PROFOUND", "Control"),
                                       labels = c("IPF (BRU)", "IPF (PROFILE)", "IPF (PROFOUND)", "Control"))
```

Before contaminant removal

```{r}
ord_nr <- ordinate(physeq_nr, method = "PCoA", distance = "wunifrac")
eig <- ord_nr$values$Eigenvalues
total_variance <- sum(eig[eig > 0])
percent_expl_pco1 <- (eig[1] / total_variance) * 100
percent_expl_pco2 <- (eig[2] / total_variance) * 100

dist_matrix <- distance(physeq_nr, method = "wunifrac")
permanova <- adonis2(dist_matrix ~ is_ctrl, data = meta(physeq_nr))
pval <- permanova$`Pr(>F)`[1]
r2 <- permanova$R2[1]

pcoa_bef <- plot_ordination(physeq_nr, ord_nr)
pcoa_bef$layers <- pcoa_bef$layers[-1]  # removes the first layer (default points)
pcoa_bef <- pcoa_bef + geom_point(aes(color = is_ctrl, shape = study), size = 0.8) + # smaller points
  stat_ellipse(aes(group = is_ctrl, color = is_ctrl, fill = is_ctrl), linetype = 2, geom = "polygon", alpha = 0.15) +
  guides(shape = guide_legend(title = "Sample Type", override.aes = list(colour = c(rep("#D1495B", 3), "#00798C"))),
         colour = "none",
         fill = "none") + 
  labs(subtitle = glue("PERMANOVA *p* = {signif(pval, 3)}<br>
                       *R*<sup>2</sup> = {signif(r2, 3)}"),
       x = glue("PCo1 ({signif(percent_expl_pco1, 3)}%)"),
       y = glue("PCo2 ({format(signif(percent_expl_pco2, 3), nsmall = 1)}%)"),
       tag = "A") +
  scale_colour_manual(values = c("#D1495B", "#00798C")) +
  scale_shape_manual(values = c(1, 2, 0, 4)) +
  scale_fill_manual(values = c("#D1495B", "#00798C")) +
  theme(plot.subtitle = element_markdown(size = 10))
```

After contaminant removal

```{r}
physeq_nc <- prune_taxa(!comb_contam$contaminant, physeq_nr)
ord_nc <- ordinate(physeq_nc, method = "PCoA", distance = "wunifrac")
eig_nc <- ord_nc$values$Eigenvalues
total_variance_nc <- sum(eig_nc[eig_nc > 0])
percent_expl_pco1_nc <- (eig_nc[1] / total_variance_nc) * 100
percent_expl_pco2_nc <- (eig_nc[2] / total_variance_nc) * 100

dist_matrix_nc <- distance(physeq_nc, method = "wunifrac")
permanova_nc <- adonis2(dist_matrix_nc ~ is_ctrl, data = meta(physeq_nc))
pval_nc <- permanova_nc$`Pr(>F)`[1]
r2_nc <- permanova_nc$R2[1]

pcoa_aft <- plot_ordination(physeq_nc, ord_nc)
pcoa_aft$layers <- pcoa_aft$layers[-1]
pcoa_aft <- pcoa_aft + geom_point(aes(color = is_ctrl, shape = study), size = 0.8) +
  stat_ellipse(aes(group = is_ctrl, color = is_ctrl, fill = is_ctrl), linetype = 2, geom = "polygon", alpha = 0.15) +
  guides(shape = guide_legend(title = "Sample Type", override.aes = list(colour = c(rep("#D1495B", 3), "#00798C"))),
         colour = "none",
         fill = "none") + 
  labs(subtitle = glue("PERMANOVA *p* = {signif(pval_nc, 3)}<br>
                       *R*<sup>2</sup> = {signif(r2_nc, 3)}"),
       x = glue("PCo1 ({signif(percent_expl_pco1_nc, 3)}%)"),
       y = glue("PCo2 ({format(signif(percent_expl_pco2_nc, 3), nsmall = 1)}%)"),
       tag = "B") +
  scale_colour_manual(values = c("#D1495B", "#00798C")) +
  scale_shape_manual(values = c(1, 2, 0, 4)) +
  scale_fill_manual(values = c("#D1495B", "#00798C")) +
  theme(plot.subtitle = element_markdown(size = 10))
```

```{r}
fig <- pcoa_bef + pcoa_aft + plot_layout(guides = "collect")
svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/contam_removal.svg", width = 12, height = 5)
print(fig)
dev.off()
```

## Alpha diversity

```{r}
physeq_alpha <- subset_samples(physeq_nc, diagnosis == "IPF" & !is.na(MUC5B))
alpha_div <- plot_richness(physeq_alpha, x = "MUC5B", color = "MUC5B", measures = c("Observed", "Shannon", "InvSimpson", "Chao1"))
alpha_div
```

### Chao1 index

Chao1 uses singletons and doubletons to estimate true genus richness. As such we apply it on the unfiltered phyloseq object.

```{r}
alpha_div_chao <- alpha_div$data %>% filter(variable == "Chao1")
chao_pval <- kruskal_test(value ~ MUC5B, data = alpha_div_chao)$p
chao_fig <- ggplot(alpha_div_chao, aes(x = MUC5B, y = value)) +
  geom_boxplot(aes(fill = MUC5B), outliers = FALSE, alpha = 0.4) +
  geom_jitter(aes(colour = MUC5B, shape = MUC5B), size = 1.2) +
  labs(subtitle = glue("Kruskal Wallis *p* = {signif(chao_pval, 3)}"),
       x = "",
       y = "Chao1 index",
       tag = "C") +
  scale_fill_manual(values = c("#2F86FF", "#EBAB16", "#DE0012"), aesthetics = c("colour", "fill")) +
  theme(plot.subtitle = element_markdown(),
        legend.position = "none")
```

Correlations with clinical variables

```{r}
alpha_div_chao <- mutate(alpha_div_chao, survival_months = floor(survival_days / 30.4375))
for (cv in clinical_vars) {
  cor_df <- cor_test(data = alpha_div_chao, value, cv, method = "spearman")
  pval <- cor_df$p
  rho <- cor_df$cor
  corr_results[str_c(cv, "chao1", sep = "_"), "p"] <- pval
  corr_results[str_c(cv, "chao1", sep = "_"), "rho"] <- rho
  if (pval < 0.05) {
    print(glue("Spearman's rho for {cv}: {rho}\np-value: {pval}"))
  }
}
```

```{r}
dlco_chao_fig <- ggplot(alpha_div_chao, aes(x = ppDLCO_baseline, y = value)) +
  geom_point(aes(colour = MUC5B, shape = MUC5B), size = 1.2) + 
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.7) +
  guides(colour = guide_legend(title = "*MUC5B* Genotype", override.aes = list(shape = c(16, 17, 15))), shape = "none") + 
  labs(subtitle = glue("Spearman's rho = {corr_results['ppDLCO_baseline_chao1', 'rho']}<br>
                       *p* < 0.0001"),
       x = "Baseline DL<sub>CO</sub> (% predicted)",
       y = "Chao1 index",
       tag = "C") +
  scale_colour_manual(values = muc5b_palette) +
  theme(plot.subtitle = element_markdown(),
        legend.title = element_markdown(face = "bold"),
        axis.title.x = element_markdown())
```

```{r}
fvc_chao_fig <- ggplot(alpha_div_chao, aes(x = ppFVC_baseline, y = value)) +
  geom_point(aes(colour = MUC5B, shape = MUC5B), size = 1.2) + 
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.7) +
  guides(colour = guide_legend(title = "*MUC5B* Genotype", override.aes = list(shape = c(16, 17, 15))), shape = "none") + 
  labs(subtitle = glue("Spearman's rho = {corr_results['ppFVC_baseline_chao1', 'rho']}<br>
                       *p* = {signif(corr_results['ppFVC_baseline_chao1', 'padj'], 3)}"),
       x = "Baseline FVC (% predicted)",
       y = "Chao1 index",
       tag = "E") +
  scale_colour_manual(values = muc5b_palette) +
  theme(plot.subtitle = element_markdown(),
        legend.title = element_markdown(face = "bold"))
```

## Remove unidentified taxa

```{r}
physeq_id <- subset_taxa(physeq_nc, !is.na(Phylum))
physeq_id
summarize_phyloseq(physeq_id)
```

## Filter out low abundance taxa (0.005% of samples or less)

```{r}
reads_per_taxa <- taxa_sums(physeq_id)
physeq_abund <- prune_taxa((reads_per_taxa / sum(reads_per_taxa)) > 0.00005, physeq_id)
physeq_abund
```

```{r}
summarize_phyloseq(physeq_abund)
```

## Normalise reads sample by sample

```{r}
normalise_reads <- function(reads) {
  reads / sum(reads)
}
physeq_rel <- transform_sample_counts(physeq_abund, normalise_reads)
```

# Export final preprocessed files

```{r}
otu_table <- as.data.frame(otu_table(physeq_rel))
tax_table <- as.data.frame(tax_table(physeq_rel))
tree <- phy_tree(physeq_rel)

write.csv(otu_table, "Preprocessed Files/otu_table.csv")
write.csv(tax_table, "Preprocessed Files/tax_table.csv")
ape::write.tree(tree, "Preprocessed Files/tree.nwk")
```

Export OTU abundance tables aggregated by different taxonomic levels (with and without metadata)

```{r}
metadata_cols <- metadata %>% rownames_to_column("sample_id") %>% 
  select(sample_id, key, diagnosis, MUC5B)

export_agg_abund <- function(physeq, level) {
  physeq_agg <- aggregate_taxa(physeq, level)
  # Combine raw reads of taxa with the same name (e.g. Prevotella and [Prevotella])
  abund_table <- as.data.frame(otu_table(physeq_agg)) %>% 
    rownames_to_column("taxon") %>% 
    mutate(taxon = str_remove_all(taxon, "\\[|\\]")) %>% # strip square brackets
    group_by(taxon) %>% 
    summarise(across(everything(), sum)) %>% # sum all columns within each group
    column_to_rownames("taxon") %>% 
    apply(2, normalise_reads) # normalise reads columnwise
  # Without metadata
  write.table(abund_table, file = glue("Preprocessed Files/{level}/{level}_rel_abund.txt"), col.names = NA, sep = "\t")
  # With metadata
  abund_table <- as.data.frame(t(abund_table)) %>% 
    rownames_to_column("sample_id") %>% 
    inner_join(metadata_cols, ., by = "sample_id")
  write_csv(abund_table, glue("Preprocessed Files/{level}/{level}_rel_abund_metadata.csv"))
}

for (level in c("Phylum", "Family", "Genus", "Species")) {
  export_agg_abund(physeq_abund, level)
}
```
