---
title: "Batch Correction"
---

# Load required packages

```{r}
library(tidyverse)
library(glue)
library(ggtext)
library(vegan)
library(MMUPHin)
```

# Load files

```{r}
setwd("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Microbiome Analysis")
meta_muc5b <- read_csv("Metadata Processing/metadata_final.csv") %>%
  drop_na(ddPCR, MUC5B) %>% filter(diagnosis == "IPF")
abund_matrix <- read_csv("Preprocessed Files/Genus/Genus_rel_abund_metadata.csv") %>% 
  drop_na(MUC5B) %>% filter(diagnosis == "IPF") %>% 
  select(-c(key, diagnosis, MUC5B)) %>% column_to_rownames(var = "sample_id")
```

# Global plot settings

```{r}
theme_set(theme_bw() + theme(legend.title = element_text(face = "bold"),
                             plot.subtitle = element_markdown(size = 10),
                             plot.tag = element_text(face = "bold")))
```

# Check for batch effect using PCoA

```{r}
set.seed(2526) # ensures reproducible results from adonis2
permanova <- adonis2(abund_matrix ~ study, data = meta_muc5b)
permanova
```

```{r}
pval <- permanova$`Pr(>F)`[1]
r2 <- permanova$R2[1]
```

```{r}
dist <- vegdist(abund_matrix)
pcoa <- cmdscale(dist, eig = TRUE)

pcoa_eig <- pcoa$eig
total_variance <- sum(pcoa_eig[pcoa_eig > 0])
percent_expl_pco1 <- (pcoa_eig[1] / total_variance) * 100
percent_expl_pco2 <- (pcoa_eig[2] / total_variance) * 100

pcoa_coord <- pcoa$points
colnames(pcoa_coord) <- c("PCo1", "PCo2")
dissimilarity <- cbind(meta_muc5b, pcoa_coord)

pcoa_bef <- ggplot(data = dissimilarity, aes(x = PCo1, y = PCo2, colour = study, shape = study)) + 
  geom_point(size = 1) + stat_ellipse(aes(fill = study), linetype = 2, geom = "polygon", alpha = 0.1) +
  guides(shape = guide_legend(title = "Cohort", override.aes = list(colour = c("#0E7175", "#FD7901", "#C35BCA"))),
         colour = "none",
         fill = "none") +
  labs(subtitle = glue("PERMANOVA *p* = {signif(pval, 3)}<br>
                       *R*<sup>2</sup> = {signif(r2, 3)}"),
       x = glue("PCo1 ({signif(percent_expl_pco1, 3)}%)"),
       y = glue("PCo2 ({signif(percent_expl_pco2, 3)}%)"),
       tag = "A") +
  scale_colour_manual(values = c("#0E7175", "#FD7901", "#C35BCA")) +
  scale_shape_manual(values = c(1, 2, 0)) +
  scale_fill_manual(values = c("#0E7175", "#FD7901", "#C35BCA"))
```

# Format input files for MMUPHin

Abundances should be a taxa by samples matrix

```{r}
abund_matrix <- t(as.matrix(abund_matrix))
```

Column names of abundance matrix should be the same as row names of metadata data frame

```{r}
meta_muc5b <- column_to_rownames(meta_muc5b, var = "sample_id")
colnames(abund_matrix) == rownames(meta_muc5b)
```

# Batch correction using MMUPHin

```{r}
fit_adjust_batch <- adjust_batch(feature_abd = abund_matrix,
                                 batch = "study",
                                 covariates = "MUC5B",
                                 data = meta_muc5b,
                                 control = list(verbose = FALSE))
abund_matrix_adj <- fit_adjust_batch$feature_abd_adj
```

# Check effect of batch correction

```{r}
abund_matrix_adj <- t(abund_matrix_adj)
permanova_adj <- adonis2(abund_matrix_adj ~ study, data = meta_muc5b)
permanova_adj
```

Smaller R^2 and F values.

```{r}
pval_adj <- permanova_adj$`Pr(>F)`[1]
r2_adj <- permanova_adj$R2[1]

dist_adj <- vegdist(abund_matrix_adj)
pcoa_adj <- cmdscale(dist_adj, eig = TRUE)

pcoa_eig_adj <- pcoa_adj$eig
total_variance_adj <- sum(pcoa_eig_adj[pcoa_eig_adj > 0])
percent_expl_pco1_adj <- (pcoa_eig_adj[1] / total_variance_adj) * 100
percent_expl_pco2_adj <- (pcoa_eig_adj[2] / total_variance_adj) * 100

pcoa_coord_adj <- pcoa_adj$points
colnames(pcoa_coord_adj) <- c("PCo1", "PCo2")
dissimilarity_adj <- cbind(meta_muc5b, pcoa_coord_adj)

pcoa_aft <- ggplot(data = dissimilarity_adj, aes(x = PCo1, y = PCo2, colour = study, shape = study)) + 
  geom_point(size = 1) + stat_ellipse(aes(fill = study), linetype = 2, geom = "polygon", alpha = 0.1) +
  guides(shape = guide_legend(title = "Cohort", override.aes = list(colour = c("#0E7175", "#FD7901", "#C35BCA"))),
         colour = "none",
         fill = "none") +
  labs(subtitle = glue("PERMANOVA *p* = {signif(pval_adj, 3)}<br>
                       *R*<sup>2</sup> = {signif(r2_adj, 3)}"),
       x = glue("PCo1 ({signif(percent_expl_pco1_adj, 3)}%)"),
       y = glue("PCo2 ({signif(percent_expl_pco2_adj, 3)}%)"),
       tag = "B") +
  scale_colour_manual(values = c("#0E7175", "#FD7901", "#C35BCA")) +
  scale_shape_manual(values = c(1, 2, 0)) +
  scale_fill_manual(values = c("#0E7175", "#FD7901", "#C35BCA"))

fig <- pcoa_bef + pcoa_aft + plot_layout(guides = "collect")
svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/batch_adjust.svg", width = 12, height = 5)
print(fig)
dev.off()
```

# Export batch corrected abundance matrix for downstream analyses

```{r}
abund_matrix_adj <- as.data.frame(abund_matrix_adj) %>% rownames_to_column("sample_id")
write_csv(abund_matrix_adj, "Preprocessed Files/Genus/Genus_rel_abund_corrected.csv")
```

# Batch correction on abundance matrix aggregated by Family

```{r}
# Load abundance matrix
abund_family <- read_csv("Preprocessed Files/Family/Family_rel_abund_metadata.csv") %>% 
  drop_na(MUC5B) %>% filter(diagnosis == "IPF") %>% 
  select(-c(key, diagnosis, MUC5B)) %>% column_to_rownames(var = "sample_id") %>% 
  as.matrix() %>% t() # taxa by samples

# Batch correction
colnames(abund_family) == rownames(meta_muc5b)
fit_adjust_batch <- adjust_batch(feature_abd = abund_family,
                                 batch = "study",
                                 covariates = "MUC5B",
                                 data = meta_muc5b,
                                 control = list(verbose = FALSE))
abund_family_adj <- fit_adjust_batch$feature_abd_adj %>% t() %>% 
  as.data.frame() %>% rownames_to_column("sample_id")

# Export corrected matrix
write_csv(abund_family_adj, "Preprocessed Files/Family/Family_rel_abund_corrected.csv")
```
