---
title: "Microbiome Analysis by MUC5B Genotype"
---

# Load required packages

```{r}
library(rstatix)
library(tidyverse)
library(reshape2)
library(vegan)
library(phyloseq)
library(microbiome)
library(glue)
library(ggrepel)
library(ggtext)
library(ggsci)
library(Maaslin2)
library(survival)
library(survminer)
library(circlize)
library(ComplexHeatmap)
library(gridtext)
library(Cairo)
library(patchwork)
```

# Load files

```{r}
setwd("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Microbiome Analysis")
meta_muc5b <- read_csv("Metadata Processing/metadata_final.csv") %>%
  drop_na(ddPCR, MUC5B) %>% filter(diagnosis == "IPF")
abund_matrix <- read_csv("Preprocessed Files/Genus/Genus_rel_abund_corrected.csv") %>%
  column_to_rownames(var = "sample_id") # batch corrected
abund_family <- read_csv("Preprocessed Files/Family/Family_rel_abund_corrected.csv") %>%
  column_to_rownames(var = "sample_id")
```

# Global plot settings

```{r}
theme_set(theme_classic() + theme(text = element_text(family = "Arial"),
                                  plot.tag = element_text(face = "bold")))
muc5b_palette <- c("#2F86FF", "#EBAB16", "#DE0012")
```

# Baseline patient characteristics

## Continuous variables

All patients

```{r}
summarise(meta_muc5b,
          age_mean = mean(age, na.rm = TRUE), age_sd = sd(age, na.rm = TRUE),
          ppFVC_mean = mean(ppFVC_baseline), ppFVC_sd = sd(ppFVC_baseline),
          ppDLCO_mean = mean(ppDLCO_baseline, na.rm = TRUE), ppDLCO_sd = sd(ppDLCO_baseline, na.rm = TRUE))
```

By *MUC5B* genotype

```{r}
group_by(meta_muc5b, MUC5B) %>% summarise(count = n(),
                                          age_mean = mean(age, na.rm = TRUE), age_sd = sd(age, na.rm = TRUE),
                                          ppFVC_mean = mean(ppFVC_baseline), ppFVC_sd = sd(ppFVC_baseline),
                                          ppDLCO_mean = mean(ppDLCO_baseline, na.rm = TRUE), ppDLCO_sd = sd(ppDLCO_baseline, na.rm = TRUE))
```

No. of samples missing age and baseline ppDLCO values

```{r}
filter(meta_muc5b, is.na(age)) %>% count(MUC5B)
```

```{r}
filter(meta_muc5b, is.na(ppDLCO_baseline)) %>% count(MUC5B)
```

Test for differences between *MUC5B* groups

```{r}
kruskal_test(age ~ MUC5B, data = meta_muc5b)
```

```{r}
kruskal_test(ppFVC_baseline ~ MUC5B, data = meta_muc5b)
```

```{r}
kruskal_test(ppDLCO_baseline ~ MUC5B, data = meta_muc5b)
```

## Categorical variables

All patients

```{r}
count(meta_muc5b, sex)
```

```{r}
count(meta_muc5b, smoking_history)
```

```{r}
count(meta_muc5b, study)
```

By *MUC5B* genotype

```{r}
count(meta_muc5b, MUC5B, sex)
```

```{r}
count(meta_muc5b, MUC5B, smoking_history)
```

```{r}
count(meta_muc5b, MUC5B, study)
```

Test for differences between *MUC5B* groups

```{r}
fisher_test(table(meta_muc5b$sex, meta_muc5b$MUC5B))
```

```{r}
fisher_test(table(meta_muc5b$smoking_history, meta_muc5b$MUC5B))
```

```{r}
fisher_test(table(meta_muc5b$study, meta_muc5b$MUC5B))
```

# Microbiome analysis

## Bacterial burden

```{r}
bact_burden_pval <- kruskal_test(ddPCR ~ MUC5B, data = meta_muc5b)$p

bb_fig <- ggplot(meta_muc5b, aes(x = MUC5B, y = ddPCR)) +
  geom_boxplot(aes(fill = MUC5B), outliers = FALSE, alpha = 0.4) +
  geom_jitter(aes(colour = MUC5B, shape = MUC5B), size = 1.2) +
  scale_y_log10(labels = scales::label_log()) + 
  labs(subtitle = glue("Kruskal Wallis *p* = {signif(bact_burden_pval, 3)}"),
       x = "",
       y = "Bacterial burden (16S rRNA gene/mL of BALF)") +
  scale_fill_manual(values = muc5b_palette, aesthetics = c("colour", "fill")) +
  theme(plot.subtitle = element_markdown(),
        legend.position = "none")

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/bact_burden.svg", width = 6, height = 5)
print(bb_fig)
dev.off()
```

## Alpha diversity

### Genus richness

```{r}
genus_no <- specnumber(abund_matrix) %>% 
  as.data.frame() %>% 
  cbind(meta_muc5b, .) %>% 
  remove_rownames() %>% 
  rename("genus_no" = ".")
```

```{r}
genus_no_pval <- kruskal_test(genus_no ~ MUC5B, data = genus_no)$p

genus_no_fig <- ggplot(genus_no, aes(x = MUC5B, y = genus_no)) +
  geom_boxplot(aes(fill = MUC5B), outliers = FALSE, alpha = 0.4) +
  geom_jitter(aes(colour = MUC5B, shape = MUC5B), size = 1.2) +
  labs(subtitle = glue("Kruskal Wallis *p* = {signif(genus_no_pval, 3)}"),
       x = "",
       y = "No. of unique genera",
       tag = "A") +
  scale_fill_manual(values = muc5b_palette, aesthetics = c("colour", "fill")) +
  theme(plot.subtitle = element_markdown(),
        legend.position = "none")
```

### Shannon index

```{r}
shannon <- vegan::diversity(abund_matrix) %>% 
  as.data.frame() %>% 
  cbind(meta_muc5b, .) %>% 
  remove_rownames() %>% 
  rename("shannon" = ".")
shannon_pval <- kruskal_test(shannon ~ MUC5B, data = shannon)$p

shannon_fig <- ggplot(shannon, aes(x = MUC5B, y = shannon)) +
  geom_boxplot(aes(fill = MUC5B), outliers = FALSE, alpha = 0.4) +
  geom_jitter(aes(colour = MUC5B, shape = MUC5B), size = 1.2) +
  labs(subtitle = glue("Kruskal Wallis *p* = {signif(shannon_pval, 3)}"),
       x = "",
       y = "Shannon index",
       tag = "B") +
  scale_fill_manual(values = muc5b_palette, aesthetics = c("colour", "fill")) +
  theme(plot.subtitle = element_markdown(),
        legend.position = "none")
```

```{r}
fig <- genus_no_fig + shannon_fig + chao_fig
svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/alpha.svg", width = 16, height = 5)
print(fig)
dev.off()
```

### Inverse Simpson index

```{r}
inv_simpson <- vegan::diversity(abund_matrix, index = "invsimpson") %>% 
  as.data.frame() %>% 
  cbind(meta_muc5b, .) %>% 
  remove_rownames() %>% 
  rename("inv_simpson" = ".")
simpson_pval <- kruskal_test(inv_simpson ~ MUC5B, data = inv_simpson)$p

ggplot(inv_simpson, aes(x = MUC5B, y = inv_simpson)) +
  geom_boxplot(aes(fill = MUC5B), outliers = FALSE, alpha = 0.4) +
  geom_jitter(aes(colour = MUC5B, shape = MUC5B), size = 1.2) +
  scale_y_continuous(breaks = seq(0, 10, 2)) +
  labs(subtitle = glue("Kruskal Wallis *p* = {signif(simpson_pval, 3)}"),
       x = "",
       y = "Inverse Simpson index") +
  scale_fill_manual(values = muc5b_palette, aesthetics = c("colour", "fill")) +
  theme(plot.subtitle = element_markdown(),
        legend.position = "none")
```

## Beta diversity

```{r}
set.seed(2526)
bray_permanova <- adonis2(abund_matrix ~ MUC5B, data = meta_muc5b)
bray_pval <- bray_permanova$`Pr(>F)`[1]
bray_r2 <- bray_permanova$R2[1]
```

Robust Aitchison is said to be better for non-negative compositional data:

```{r}
adonis2(abund_matrix ~ MUC5B, data = meta_muc5b, method = "robust.aitchison")
```

```{r}
bray_dist <- vegdist(abund_matrix)
pcoa <- cmdscale(bray_dist, eig = TRUE)

pcoa_eig <- pcoa$eig
total_variance <- sum(pcoa_eig[pcoa_eig > 0])
percent_expl_pco1 <- (pcoa_eig[1] / total_variance) * 100
percent_expl_pco2 <- (pcoa_eig[2] / total_variance) * 100

pcoa_coord <- pcoa$points
colnames(pcoa_coord) <- c("PCo1", "PCo2")
bray_dissimilarity <- cbind(meta_muc5b, pcoa_coord)

beta_fig <- ggplot(data = bray_dissimilarity, aes(x = PCo1, y = PCo2, colour = MUC5B)) + 
  geom_point(aes(shape = MUC5B), size = 1.5) + stat_ellipse(aes(fill = MUC5B), linetype = 2, geom = "polygon", alpha = 0.1) +
  guides(shape = guide_legend(title = "*MUC5B* Genotype", override.aes = list(colour = muc5b_palette)),
         colour = "none",
         fill = "none") + 
  labs(subtitle = glue("PERMANOVA *p* = {signif(bray_pval, 3)}<br>
                       R<sup>2</sup> = {signif(bray_r2, 3)}"),
       x = glue("PCo1 ({signif(percent_expl_pco1, 3)}%)"),
       y = glue("PCo2 ({signif(percent_expl_pco2, 3)}%)")) +
  scale_colour_manual(values = muc5b_palette) +
  scale_fill_manual(values = muc5b_palette) +
  theme_bw() + theme(plot.subtitle = element_markdown(size = 10),
                     legend.title = element_markdown(face = "bold"))

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/beta.svg", width = 7, height = 5)
print(beta_fig)
dev.off()
```

## Top 10 taxa

### By Genus

Identify top 10 genera

```{r}
abund_matrix <- abund_matrix * 100 # convert to %
rowSums(abund_matrix)
abund_sorted <- abund_matrix[, order(colSums(abund_matrix), decreasing = TRUE)]
top_ten <- colnames(abund_sorted)[1:10]
top_ten
```

Select abundance counts of top 10 genera and combine with metadata

```{r}
top_ten_abund <- cbind(meta_muc5b, abund_sorted[, top_ten]) %>% remove_rownames()
```

Test for statistically significant differences between *MUC5B* genotypes

```{r}
for (genus in top_ten) {
  kruskal_result <- kruskal_test(top_ten_abund, as.formula(glue("{genus} ~ MUC5B")))
  print(kruskal_result)
  if (kruskal_result$p < 0.05) {
    # Post-hoc Dunn's test
    print(dunn_test(top_ten_abund, as.formula(glue("{genus} ~ MUC5B")), p.adjust.method = "holm"))
  }
}
```

```{r}
top_ten_stats <- top_ten_abund %>%
  select(sample_id, MUC5B, Streptococcus:Actinobacillus) %>% 
  melt(id.vars = "MUC5B", # reshape to long data format
       measure.vars = top_ten, 
       variable.name = "Genus",
       factorsAsStrings = TRUE) %>% # Genus is converted to a factor
  group_by(MUC5B, Genus) %>%
  summarise(median = median(value),
            mean = mean(value),
            q1 = quantile(value, 0.25),
            q3 = quantile(value, 0.75)) %>% 
  mutate(Phylum = case_when(Genus == "Rothia" ~ "Actinobacteria",
                            Genus == "Actinomyces" ~ "Actinomycetota",
                            Genus %in% c("Gemella", "Granulicatella") ~ "Bacillota",
                            Genus == "Prevotella" ~ "Bacteroidetes",
                            Genus %in% c("Streptococcus", "Veillonella") ~ "Firmicutes",
                            Genus %in% c("Actinobacillus", "Haemophilus", "Neisseria") ~ "Pseudomonadota"))
```

Horizontal bar chart

```{r}
tt_fig <- ggplot(top_ten_stats, aes(x = median, y = Genus)) +
  geom_col(aes(fill = Phylum), colour = "black", linewidth = 0.5) +
  geom_errorbar(aes(xmin = q1, xmax = q3, y = Genus), orientation = "y",
                width = 0.3, linewidth = 0.5) +
  labs(x = "Relative abundance (%)", y = "") +
  facet_grid(. ~ MUC5B) +
  scale_fill_brewer(palette = "Dark2") +
  theme(axis.text.y = element_text(size = 11, face = "italic", colour = c(rep("#852F88", 3), rep("black", 7))),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        strip.background = element_blank(), # remove boxes around facet titles
        strip.text = element_text(size = 14, face = "bold"))

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/top_ten.svg", width = 8, height = 5)
print(tt_fig)
dev.off()
```

Vertical column chart

```{r}
ggplot(top_ten_stats, aes(x = Genus, y = median)) +
  geom_col(aes(fill = Phylum), colour = "black", linewidth = 0.5) +
  geom_errorbar(aes(x = Genus, ymin = q1, ymax = q3),
                width = 0.3, linewidth = 0.5) +
  labs(x = "", y = "Relative abundance (%)") +
  facet_grid(MUC5B ~ .) +
  scale_fill_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 11, face = "italic"),
        legend.title = element_text(hjust = 0.5, face = "bold"),
        legend.text = element_text(face = "italic"),
        strip.background = element_blank(),
        strip.text.y.right = element_text(angle = 0, size = 14, face = "bold"))
```

### By Family

```{r}
# Top 10 families
abund_family <- abund_family * 100
rowSums(abund_family)
abund_family_sorted <- abund_family[, order(colSums(abund_family), decreasing = TRUE)]
top_ten_families <- colnames(abund_family_sorted)[1:10]
top_ten_families

# Abundance counts of top 10 families
top_ten_abund_family <- cbind(meta_muc5b, abund_family_sorted[, top_ten_families]) %>% remove_rownames()

# Test for differences between MUC5B genotypes
for (family in top_ten_families) {
  kruskal_result <- kruskal_test(top_ten_abund_family, as.formula(glue("{family} ~ MUC5B")))
  print(kruskal_result)
  if (kruskal_result$p < 0.05) {
    # Post-hoc Dunn's test
    print(dunn_test(top_ten_abund_family, as.formula(glue("{family} ~ MUC5B")), p.adjust.method = "holm"))
  }
}
```

Bar chart

```{r}
top_ten_stats_family <- top_ten_abund_family %>%
  select(sample_id, MUC5B, Streptococcaceae:Staphylococcaceae) %>% 
  melt(id.vars = "MUC5B", # reshape to long data format
       measure.vars = top_ten_families, 
       variable.name = "Family",
       factorsAsStrings = TRUE) %>% # Family is converted to a factor
  group_by(MUC5B, Family) %>%
  summarise(median = median(value),
            mean = mean(value),
            q1 = quantile(value, 0.25),
            q3 = quantile(value, 0.75))

ggplot(top_ten_stats_family, aes(x = median, y = Family)) +
  geom_col(aes(fill = Family), colour = "black", linewidth = 0.5) +
  geom_errorbar(aes(xmin = q1, xmax = q3, y = Family), orientation = "y",
                width = 0.3, linewidth = 0.5) +
  labs(x = "Relative abundance (%)", y = "") +
  facet_grid(. ~ MUC5B) +
  theme(axis.text.y = element_text(size = 11, face = "italic"),
        # legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        strip.background = element_blank(), # remove boxes around facet titles
        strip.text = element_text(size = 14, face = "bold"))
```

## Top 50 taxa

### By Genus

```{r}
# Top 50 genera
top_fifty <- colnames(abund_sorted)[c(1:13, 15:51)] # 14th genus is Unknown
top_fifty

# Abundance counts of top 50 genera
top_fifty_abund <- cbind(meta_muc5b, abund_sorted[, top_fifty]) %>% remove_rownames()

# Test for differences between MUC5B genotypes
top_genus_results <- data.frame(p = rep(0, times = 50), padj = rep(0, times = 50), row.names = top_fifty)

for (genus in top_fifty) {
  kruskal_result <- kruskal_test(top_fifty_abund, as.formula(glue("{genus} ~ MUC5B")))
  top_genus_results[genus, "p"] <- kruskal_result$p
}

top_genus_results$padj <- p.adjust(top_genus_results$p, method = "BH")
top_genus_results$padj < 0.05
```

### By Family

```{r}
# Top 50 families
top_fifty_families <- colnames(abund_family_sorted)[c(1:18, 20:49)] # 19th family is Unknown, all other families are not present in any sample
top_fifty_families

# Abundance counts of top 50 families
top_fifty_abund_family <- cbind(meta_muc5b, abund_family_sorted[, top_fifty_families]) %>% remove_rownames()

# Test for differences between MUC5B genotypes
top_family_results <- data.frame(p = rep(0, times = 48), padj = rep(0, times = 48), row.names = top_fifty_families)

for (family in top_fifty_families) {
  kruskal_result <- kruskal_test(top_fifty_abund_family, as.formula(glue("{family} ~ MUC5B")))
  top_family_results[family, "p"] <- kruskal_result$p
}

top_family_results$padj <- p.adjust(top_family_results$p, method = "BH")
top_family_results$padj < 0.05
```

## Differentially abundant taxa

### By Genus

```{r}
meta_muc5b <- column_to_rownames(meta_muc5b, "sample_id")
Maaslin2(abund_matrix, meta_muc5b, output = "MaAsLin2 Output/Genus",
         min_prevalence = 0, normalization = "NONE",
         # abundance data is already pre-filtered and pre-normalised
         fixed_effects = c("MUC5B"), reference = c("MUC5B,G/G"))
```

*Bacteroides* is significantly differentially expressed between G/G and T/T groups.

```{r}
bacteroides_abund <- cbind(select(meta_muc5b, MUC5B), abund_matrix$Bacteroides) %>%
  rename(Bacteroides = "abund_matrix$Bacteroides")
summary(bacteroides_abund)
```

Only 1 patient with T/T genotype has *Bacteroides* found in their microbiome. *Bacteroides* is absent from all other patients.

### By Family

```{r}
Maaslin2(abund_family, meta_muc5b, output = "MaAsLin2 Output/Family",
         min_prevalence = 0.1, normalization = "NONE",
         fixed_effects = c("MUC5B"), reference = c("MUC5B,G/G"))
```

# Survival analysis

```{r}
meta_muc5b <- mutate(meta_muc5b, survival_months = floor(survival_days / 30.4375))
meta_muc5b <- mutate(meta_muc5b, death = if_else(death == "Yes", 1, 0))
```

## Fit survival data using Kaplan-Meier method

```{r}
surv_object <- Surv(time = meta_muc5b$survival_months, event = meta_muc5b$death)
surv_object
surv_muc5b <- survfit(surv_object ~ MUC5B, data = meta_muc5b)
summary(surv_muc5b)
```

```{r}
kaplan_pval <- surv_pvalue(surv_muc5b, data = meta_muc5b)$pval
```

```{r}
survplot <- ggsurvplot(surv_muc5b, data = meta_muc5b, fun = "pct",
                       palette = muc5b_palette, size = 0.5,
                       conf.int = TRUE, conf.int.alpha = 0.15, risk.table = TRUE,
                       xlab = "Time (months)",
                       legend = "right", legend.title = "*MUC5B* Genotype",
                       legend.labs = c("G/G", "G/T", "T/T"),
                       break.time.by = 6, xlim = c(0, 36), ylim = c(40, 100),
                       font.main = 11, font.x = 11, font.y = 11,
                       font.tickslab = 11, font.legend = 11,
                       risk.table.fontsize = 4) # risk table uses a different scale
survplot$plot <- survplot$plot +
  geom_richtext(aes(x = 1, y = 50, label = glue("Log-rank *p* = {signif(kaplan_pval, 3)}")),
                fill = NA, label.color = NA, hjust = 0, size = 4) +
  theme(legend.title = element_markdown(face = "bold"))
survplot$table <- survplot$table +
  theme(axis.line.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size = 11))
survplot

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/kaplan_meier.svg", width = 8, height = 5)
print(survplot)
dev.off()
```

No significant difference in survival between groups but T/T group has a gentler gradient, possibly indicating better survival.

## Fit Cox proportional hazards model

```{r}
coxph_muc5b <- coxph(surv_object ~ MUC5B + study, data = meta_muc5b)
ggforest(coxph_muc5b, data = as.data.frame(meta_muc5b))
```

# Clustered heatmap

## Cluster samples

```{r}
top_ten_counts <- abund_sorted[, top_ten]
which(rowSums(top_ten_counts) == 0)
```

Remove BRU.05380 from heatmap

```{r}
top_ten_counts <- filter(top_ten_counts, row.names(top_ten_counts) != "BRU.05380")
```

UPGMA on Bray-Curtis dissimilarities

```{r}
sample_dist <- vegdist(top_ten_counts)
col_clustering <- hclust(sample_dist, "average")
genera_dist <- vegdist(t(top_ten_counts))
row_clustering <- hclust(genera_dist, "average")
```

## Plot heatmap

```{r}
heatmap_matrix <- as.matrix(t(top_ten_counts)) # taxa as rows, samples as columns
```

Create column annotations

```{r}
anno_df <- meta_muc5b %>% filter(row.names(meta_muc5b) != "BRU.05380") %>% 
  select(c(FVC_change_1y, status_1y, MUC5B))

rownames(anno_df) == colnames(heatmap_matrix) # check that samples are in the same order

fvc_palette <- colorRamp2(seq(-50, 50, 25), c("#B2182B", "#F4A582", "white", "#92C5DE", "#2166AC"))
muc5b_palette_anno <- muc5b_palette
names(muc5b_palette_anno) <- c("G/G", "G/T", "T/T")
anno_palette <- list(FVC_change_1y = fvc_palette,
                     status_1y = c("Stable" = "#499F31",
                                   "Semi-progressive" = "#B896D6",
                                   "Progressive" = "#783DBA"),
                     MUC5B = muc5b_palette_anno)

col_anno <- HeatmapAnnotation(df = anno_df,
                              col = anno_palette,
                              annotation_label = gt_render(c("Relative FVC Change (%)",
                                                             "Status",
                                                             "*MUC5B* Genotype")),
                              annotation_name_gp = gpar(fontsize = 10),
                              annotation_legend_param = list(FVC_change_1y = list(title = "Relative\nFVC Change (%)"),
                                                             status_1y = list(title = "Status"),
                                                             MUC5B = list(title = gt_render("*MUC5B* Genotype"))))
```

Option to arrange columns by *MUC5B* genotype

```{r}
anno_df_by_geno <- arrange(anno_df, MUC5B)
rownames(anno_df_by_geno)
```

Create heatmap with column annotations

```{r}
abund_palette <- colorRamp2(seq(0, 100, 20), c("white", "#F8D7DA", "#EA868F", "#DC3545", "#842029", "#2C0B0E"))

hm1 <- Heatmap(heatmap_matrix,
        name = "Relative\nAbundance (%)", # legend title
        col = abund_palette,
        rect_gp = gpar(col = "white", lwd = 0.5), # cell borders
        
        column_title = "Samples",
        column_title_side = "bottom",
        column_title_gp = gpar(fontsize = 10),
        
        cluster_rows = row_clustering, # previously calculated clustering
        cluster_columns = col_clustering,
        
        row_names_side = "left",
        row_names_gp = gpar(fontsize = 10, fontface = "italic"),
        show_column_names = FALSE,
        
        top_annotation = col_anno, # column annotations
        
        heatmap_legend_param = list(title_gp = gpar(fontsize = 10, fontface = "bold"),
                                    at = seq(0, 100, 20),
                                    labels = as.character(seq(0, 100, 20)),
                                    labels_gp = gpar(fontsize = 10)))

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/heatmap.svg", width = 15, height = 6)
draw(hm1)
dev.off()

hm2 <- Heatmap(heatmap_matrix,
        name = "Relative\nAbundance (%)",
        col = abund_palette,
        rect_gp = gpar(col = "white", lwd = 0.5),
        
        column_title = "Samples",
        column_title_side = "bottom",
        column_title_gp = gpar(fontsize = 10),
        
        cluster_rows = row_clustering,
        column_order = rownames(anno_df_by_geno),
        
        row_names_side = "left",
        row_names_gp = gpar(fontsize = 10, fontface = "italic"),
        show_column_names = FALSE,
        
        top_annotation = col_anno,
        
        heatmap_legend_param = list(title_gp = gpar(fontsize = 10, fontface = "bold"),
                                    at = seq(0, 100, 20),
                                    labels = as.character(seq(0, 100, 20)),
                                    labels_gp = gpar(fontsize = 10)))

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/heatmap_by_muc5b.svg", width = 15, height = 6)
draw(hm2)
dev.off()
```

## Test for associations between microbial composition and clinical variables

```{r}
adonis2(abund_matrix ~ FVC_change_1y, data = meta_muc5b, na.action = na.omit)
```

```{r}
adonis2(abund_matrix ~ status_1y, data = meta_muc5b, na.action = na.omit)
```

# Microbiome correlations with clinical variables

```{r}
clinical_vars <- c("ppFVC_baseline", "ppDLCO_baseline", "FVC_change_1y")

corr_results <- data.frame(p = rep(0, 12), rho = rep(0, 12), padj = rep(0, 12),
                           row.names = c("ppFVC_baseline_burden", "ppDLCO_baseline_burden", "FVC_change_1y_burden",
                                         "ppFVC_baseline_richness", "ppDLCO_baseline_richness", "FVC_change_1y_richness",
                                         "ppFVC_baseline_shannon", "ppDLCO_baseline_shannon", "FVC_change_1y_shannon",
                                         "ppFVC_baseline_chao1", "ppDLCO_baseline_chao1", "FVC_change_1y_chao1"))
```

Bacterial burden

```{r}
for (cv in clinical_vars) {
  cor_df <- cor_test(data = meta_muc5b, ddPCR, cv, method = "spearman")
  pval <- cor_df$p
  rho <- cor_df$cor
  corr_results[str_c(cv, "burden", sep = "_"), "p"] <- pval
  corr_results[str_c(cv, "burden", sep = "_"), "rho"] <- rho
  if (pval < 0.05) {
    print(glue("Spearman's rho for {cv}: {rho}\np-value: {pval}"))
  }
}
```

Genus richness

```{r}
for (cv in clinical_vars) {
  cor_df <- cor_test(data = genus_no, genus_no, cv, method = "spearman")
  pval <- cor_df$p
  rho <- cor_df$cor
  corr_results[str_c(cv, "richness", sep = "_"), "p"] <- pval
  corr_results[str_c(cv, "richness", sep = "_"), "rho"] <- rho
  if (pval < 0.05) {
    print(glue("Spearman's rho for {cv}: {rho}\np-value: {pval}"))
  }
}
```

Shannon index

```{r}
for (cv in clinical_vars) {
  cor_df <- cor_test(data = shannon, shannon, cv, method = "spearman")
  pval <- cor_df$p
  rho <- cor_df$cor
  corr_results[str_c(cv, "shannon", sep = "_"), "p"] <- pval
  corr_results[str_c(cv, "shannon", sep = "_"), "rho"] <- rho
  if (pval < 0.05) {
    print(glue("Spearman's rho for {cv}: {rho}\np-value: {pval}"))
  }
}
```

Adjust *p*-values for multiple comparisons

```{r}
corr_results$padj <- p.adjust(corr_results$p, method = "BH")
corr_results$is_signif <- corr_results$padj < 0.05
```

## Linear regression plots

Genus richness against ppDLCO

```{r}
dlco_gen_fig <- ggplot(genus_no, aes(x = ppDLCO_baseline, y = genus_no)) +
  geom_point(aes(colour = MUC5B, shape = MUC5B), size = 1.2) + 
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.7) +
  guides(colour = guide_legend(title = "*MUC5B* Genotype", override.aes = list(shape = c(16, 17, 15))),
         shape = "none") + 
  labs(subtitle = glue("Spearman's rho = {corr_results['ppDLCO_baseline_richness', 'rho']}<br>
                       *p* = {signif(corr_results['ppDLCO_baseline_richness', 'padj'], 3)}"),
       x = "Baseline DL<sub>CO</sub> (% predicted)",
       y = "No. of unique genera",
       tag = "A") +
  scale_colour_manual(values = muc5b_palette) +
  theme(plot.subtitle = element_markdown(),
        legend.title = element_markdown(face = "bold"),
        axis.title.x = element_markdown())
```

Genus richness against ppFVC

```{r}
fvc_gen_fig <- ggplot(genus_no, aes(x = ppFVC_baseline, y = genus_no)) +
  geom_point(aes(colour = MUC5B, shape = MUC5B), size = 1.2) + 
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.7) +
  guides(colour = guide_legend(title = "*MUC5B* Genotype", override.aes = list(shape = c(16, 17, 15))),
         shape = "none") + 
  labs(subtitle = glue("Spearman's rho = {corr_results['ppFVC_baseline_richness', 'rho']}<br>
                       *p* = {signif(corr_results['ppFVC_baseline_richness', 'padj'], 3)}"),
       x = "Baseline FVC (% predicted)",
       y = "No. of unique genera",
       tag = "D") +
  scale_colour_manual(values = muc5b_palette) +
  theme(plot.subtitle = element_markdown(),
        legend.title = element_markdown(face = "bold"))
```

Shannon index against ppDLCO

```{r}
dlco_shannon_fig <- ggplot(shannon, aes(x = ppDLCO_baseline, y = shannon)) +
  geom_point(aes(colour = MUC5B, shape = MUC5B), size = 1.2) + 
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.7) +
  guides(colour = guide_legend(title = "*MUC5B* Genotype", override.aes = list(shape = c(16, 17, 15))),
         shape = "none") + 
  labs(subtitle = glue("Spearman's rho = {corr_results['ppDLCO_baseline_shannon', 'rho']}<br>
                       *p* = {signif(corr_results['ppDLCO_baseline_shannon', 'padj'], 3)}"),
       x = "Baseline DL<sub>CO</sub> (% predicted)",
       y = "Shannon index",
       tag = "B") +
  scale_colour_manual(values = muc5b_palette) +
  theme(plot.subtitle = element_markdown(),
        legend.title = element_markdown(face = "bold"),
        axis.title.x = element_markdown())
```

```{r}
fig <- (dlco_gen_fig | dlco_shannon_fig | dlco_chao_fig) /
  (plot_spacer() | fvc_gen_fig | fvc_chao_fig | plot_spacer()) +
  plot_layout(guides = "collect", widths = c(1, 1, 1))

svg("~/OneDrive - Imperial College London/MSc Genomic Medicine/Research Project/Dissertation/Figures/clin_feat.svg", width = 18, height = 10)
print(fig)
dev.off()
```
